Recursive Solutions to First-Order Model Counting

### 2022-02-25 notes/notes.tex
A table to summarise experimental results

\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{amsfonts}
\usepackage{amssymb}

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}

@@

\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}

@@

@@

\paragraph{Things I might need to explain.}
\begin{itemize}
\item atom, (positive/negative) literal, constant, predicate, variable, literal variable, clause, unit clause
\item $\Vars$, $\Vars(c) = \Vars(L) \cup \Vars(C)$
\item $\Doms$ on both formulas and clauses. $\Doms(c) = \Imm \delta_c$, and $\Doms(\phi) = \bigcup_{c \in \phi} \Doms(c)$.
\item $\WMC$, $\wwp$, $\wwn$, $\Imm$
\item two parts: compilation and inference.
\item During inference, there is a domain size map $\size\colon \mathcal{D} \to \mathbb{N}_0$.
\item mention which rules are in $\Gamma$ and which ones are in $\Delta$ (and why tryCache has to be in $\Delta$).
\item \textsc{ForcLift}
\item In a way, we're dividing the idea of domain recursion between the IDR and the Ref nodes, thus also generalising it. [This is good context to refer to \cref{fig:idea}.]
\end{itemize}

@@

\paragraph{TODO}
\begin{itemize}
\item capitalise variable names.
\end{itemize}

@@

@@

\subsection{From Circuits to Labelled Graphs}

@@

@@

\paragraph{TODO.}
\begin{itemize}
\item provide a short explanation of the types (emphasising which ones are new/updated). %\footnote{The new (as well as significantly renewed) types in \mathcal{T} are described in subsequent (TODO: which ones?) sections. The meanings of other types are described...}
\item have an example of a simple FCG
\end{itemize}

\subsection{Everything Else}

@@

\section{Smoothing}

@@

\section{Identifying Possibilities for Recursion}

@@

\subsection{Evaluation}

@@

Throughout this section, let $\phi$ be an arbitrary formula.

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\DR$}
  \label{alg:domainrecursion}
  \KwIn{formula $\phi$}
  \KwOut{set $S$}
  $S \gets \emptyset$\;
  \ForEach{domain $d \in \mathcal{D}$ s.t. there is a clause $c \in \phi$ and a variable $v \in \Vars(L_c)$ s.t. $\delta_c(v) = d$}{
    $\phi' \gets \emptyset$\;
    $x \gets \text{a new constant associated with domain } d$\;
    \ForEach{clause $c = (L, C, \delta) \in \phi$\label{line:forclause}}{
      $V \gets \{\, v \in \Vars(L) \mid \delta(v) = d \,\}$\;
      \ForEach{subset $W \subseteq V$ s.t. $W^2 \cap C = \emptyset$ {\bf and} $W \cap \{\, v \in \Vars(C) \mid (v, y) \in C \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
        \tcc{Here, $\delta'$ is the restriction of $\delta$ to the new set of variables}
        $\phi' \gets \phi' \cup \{\, (L[x/W], C[x/W] \cup \{\, (v, x) \mid (v \in V \setminus W) \,\}, \delta') \,\}$\;
      }
    }
    $S \gets S \cup \{\, (\DR_d, \phi') \,\}$\;
  }
\end{algorithm}

@@

\paragraph{Evaluation.}
\[
\WMC(\DR_d(n); \size) =
\begin{cases}
  \WMC(n; \size) & \text{if } \size(d) > 0 \\
  1 & \text{otherwise.}
\end{cases}
\]

@@

\section{Newly Domain-Liftable Formulas}

\begin{figure}
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=48mm]
    \node (dr) {$\DR_{x \in a}$}
    child {node {$\bigvee_{b^\top \subseteq b}$}
      child {node {$\wedge$}
        child {node {$1_{(\{\, p(x, X) \,\}, \emptyset, \{\, X \mapsto b^\top \,\})}$}}
        child {node {$\wedge$}
          child {node {$1_{(\{\, \neg p(x, X) \,\}, \emptyset, \{\, X \mapsto b^\bot \,\})}$}}
          child {node {$\wedge$}
            child {node {$1_{(\{\, \neg p(X, Y) \,\}, \{\, (X, x) \,\}, \{\, X \mapsto a, Y \mapsto b^\top \,\})}$}}
            child {node {$\CR_{a \mapsto a'}$}
              child {node {$\wedge$}
                child {node {$\bot_{(\emptyset, \{\, (X, Y) \,\}, \{\, X \mapsto b^\top, Y \mapsto b^\top \,\})}$}}
                child {node (ref) {$\Reff_{\{\, a \mapsto a', b \mapsto b^\bot \,\}}$}}
              }}}}}};
    \draw[-latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A graphical representation of an FCG for injections and partial injections between two domains. TODO: add a reference to a formula?}
\end{figure}

\paragraph{TODO.}
\begin{itemize}
\item $b^\bot$ should be part of the notation for counting.
\item \textsc{ForcLift} fails on all of these.
\item Mention that functions, surjections, and their partial counterparts are/were already liftable.
\item Formulas for partial injections, how to add totality, surjectivity.
\item Describe the figure and connect it to the algebraic formula.
\item Explain the algebraic notation that I'm using here (e.g., that $f$ is always the main function)
\item Explain the Knuth's bracket notation (or use something else, since I'm already using brackets for lists)
\item Explain the importance of comparing domain sizes to 2.
\item Maybe a big table: name, formula, optimal solution and its complexity, best found solution, its depth, its complexity. Also mention that it only takes a few seconds to find these solutions.
\item Going beyond depth 6 (or sometimes even completing depth 6) is computationally infeasible with the current implementation, but depth at most 5 can be searched within at most a few seconds.
\item Notation for domain recursion should include the constant name.
\item Combine the tikz and the algebraic notation into one, so I don't need to have two versions. But how? Maybe associate a symbol with each type and only to the types that I use?
\item Partial bijection is the same as partial injection.
\item the exponential solutions can be computed in quadratic time with dynamic programming!
\end{itemize}

\begin{remark}
  FCGs that compare the size of a domain to an integer can be constructed automatically using compilation rules, although $n$ is upper bounded by the maximum number of variables in any clause of the input formula since there is no rule that would introduce new variables.
\end{remark}

\begin{itemize}
\item 1d bijections and 1d injections (note that it's the same problem). Depth 3 solution:
  \begin{align*}
    f(n) &= \sum_{m=0}^n \binom{n}{m} (-1)^{n-m}g(n, m) \\
    g(n, m) &= \sum_{l=0}^n \binom{n}{l}[l < 2]g(n-l, m-1) \\
    &= g(n, m - 1) + ng(n - 1, m - 1),
  \end{align*}
  which works with base case $g(n, 0) = 1$.
\item 1d partial injections. 2 solutions at depth 6, but they're too complicated to check by hand. A contradiction with $X \ne x$ constraints makes things complicated.
\item 2d bijections. Depth 3:
  \begin{align*}
    f(m, n) &= \sum_{l=0}^m \binom{m}{l} [l < 2] (1 - [l < 1])f(m-l, n-1) \\
    &= mf(m-1, n-1),
  \end{align*}
  which works with base cases $f(0, 0) = 1$, $f(0, n) = 0$, $f(m, 0) = 0$.
\item 2d injections. Depth 2:
  \begin{align*}
    f(m, n) &= \sum_{l=0}^m \binom{m}{l}[l<2]f(m-l, n-1) \\
    &= f(m, n-1) + mf(m-1, n-1),
  \end{align*}
  which works with base cases $f(0, 0) = 1$ and $f(m, 0) = 0$.
\item 2d partial injections, depth 2. Exactly the same circuit as above but with base case $f(m, 0) = 1$.
\end{itemize}

>>>>

\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{pifont}

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{colortbl}

@@

\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

@@

\section{Introduction}

@@

\paragraph{Things I might need to explain.}
\begin{itemize}
\item atom, (positive/negative) literal, constant, predicate, argument, arity, variable, clause, unit clause, first-order logic
\item $\Vars$, $\Vars(c) = \Vars(L) \cup \Vars(C)$
\item $\Doms$ on both formulas and clauses. $\Doms(c) = \Imm \delta_c$, and $\Doms(\phi) = \bigcup_{c \in \phi} \Doms(c)$.
\item $\WMC$, $\wwp$, $\wwn$, $\Imm$
\item two parts: compilation and inference.
\item During inference, there is a domain size map $\size\colon \mathcal{D} \to \mathbb{N}_0$.
\item mention which rules are in $\Gamma$ and which ones are in $\Delta$ (and why tryCache has to be in $\Delta$).
\item \textsc{ForcLift}
\item In a way, we're dividing the idea of domain recursion between the IDR and the Ref nodes, thus also generalising it. [This is good context to refer to \cref{fig:idea}.]
\end{itemize}

@@

@@

TODO: implicit universal quantification.

@@

\section{From Circuits to Labelled Graphs}

@@

TODO: conjunctions and disjunctions should be different from set-conjunctions and set-disjunctions.

@@

\paragraph{TODO.}
\begin{itemize}
\item provide a short explanation of the types (emphasising which ones are new/updated). %\footnote{The new (as well as significantly renewed) types in \mathcal{T} are described in subsequent (TODO: which ones?) sections. The meanings of other types are described...}
\item have an example of a simple FCG. Describe the figure from later.
\end{itemize}

@@

\section{In-Between Compilation and Inference: Smoothing}

@@

\subsection{Identifying Possibilities for Recursion}

@@

\subsubsection{Evaluation}

@@

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\DR$}
  \label{alg:domainrecursion}
  \KwIn{formula $\phi$}
  \KwOut{set $S$}
  $S \gets \emptyset$\;
  \ForEach{domain $d \in \mathcal{D}$ s.t. there is a clause $c \in \phi$ and a variable $v \in \Vars(L_c)$ s.t. $\delta_c(v) = d$}{
    $\phi' \gets \emptyset$\;
    $x \gets \text{a new constant associated with domain } d$\;
    \ForEach{clause $c = (L, C, \delta) \in \phi$\label{line:forclause}}{
      $V \gets \{\, v \in \Vars(L) \mid \delta(v) = d \,\}$\;
      \ForEach{subset $W \subseteq V$ s.t. $W^2 \cap C = \emptyset$ {\bf and} $W \cap \{\, v \in \Vars(C) \mid (v, y) \in C \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
        \tcc{Here, $\delta'$ is the restriction of $\delta$ to the new set of variables}
        $\phi' \gets \phi' \cup \{\, (L[x/W], C[x/W] \cup \{\, (v, x) \mid (v \in V \setminus W) \,\}, \delta') \,\}$\;
      }
    }
    $S \gets S \cup \{\, (\DR_{d \gets d \setminus \{\, x \,\}}, \phi') \,\}$\;
  }
\end{algorithm}

@@

\paragraph{Evaluation.}
\[
\WMC(\DR_{d \gets d \setminus \{\, x \,\}}(n); \size) =
\begin{cases}
  \WMC(n; \size) & \text{if } \size(d) > 0 \\
  1 & \text{otherwise.}
\end{cases}
\]

@@

\section{Examples of Newly Domain-Liftable Formulas}

\begin{table}
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Complexity} \\
    Partial & Endo & Class & Best Known & With Circuits & With Graphs \\
    \midrule
    \rowcolor{lightgray}\cmark/\xmark & \cmark/\xmark & Functions & $\lg m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \lg m$ \cite{30049} & ? & ? \\
    \xmark & \cmark & & $n \lg m$ \cite{30049} & ? & ? \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $b$ to $a$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{lightgray}\xmark & \xmark & & $\lg m$ & - & $mn$ \\
    \rowcolor{lightgray}\xmark & \cmark & & $\lg m$ & - & $m^3$ \\
    \rowcolor{lightgray}\cmark & \xmark & & $\min\{\, m, n \,\}^2$ & - & $n^2$ \\
    \rowcolor{lightgray}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & - & - \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $\lg m$ & - & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{Here, $m$ is the size of $a$, and $n$ is the size of $b$. All asymptotic complexities are in $\Theta(\cdot)$. Logarithm with base two. This is for unweighted counting. A hyphen means that no solution was found. TODO: add shaded rows. TODO: explain assumptions for the counts to not be zero. TODO: double check. TODO: maybe combine more rows/columns.}
\end{table}

\begin{figure}
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=48mm]
    \node (dr) {$\DR_{a \gets a \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{b^\top \subseteq b}$}
      child {node {$\wedge$}
        child {node {$1_{(\{\, p(x, X) \,\}, \emptyset, \{\, X \mapsto b^\top \,\})}$}}
        child {node {$\wedge$}
          child {node {$1_{(\{\, \neg p(x, X) \,\}, \emptyset, \{\, X \mapsto b^\bot \,\})}$}}
          child {node {$\wedge$}
            child {node {$1_{(\{\, \neg p(X, Y) \,\}, \{\, (X, x) \,\}, \{\, X \mapsto a, Y \mapsto b^\top \,\})}$}}
            child {node {$\CR_{a \mapsto a'}$}
              child {node {$\wedge$}
                child {node {$\bot_{(\emptyset, \{\, (X, Y) \,\}, \{\, X \mapsto b^\top, Y \mapsto b^\top \,\})}$}}
                child {node (ref) {$\Reff_{\{\, a \mapsto a', b \mapsto b^\bot \,\}}$}}
              }}}}}};
    \draw[-latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A graphical representation of an FCG for injections and partial injections between two domains. TODO: add a reference to a formula?}
  \label{fig:examplefcg}
\end{figure}

\paragraph{TODO.}
\begin{itemize}
\item $b^\bot$ should be part of the notation for counting.
\item Formulas for partial injections, how to add totality, surjectivity.
\item Describe \cref{fig:examplefcg} and connect it to the algebraic formula.
\item Explain the algebraic notation that I'm using here (e.g., that $f$ is always the main function)
\item Explain the importance of comparing domain sizes to 2. FCGs that compare the size of a domain to an integer can be constructed automatically using compilation rules, although $n$ is upper bounded by the maximum number of variables in any clause of the input formula since there is no rule that would introduce new variables.
\item Maybe a big table: name, optimal solution and its complexity, best found solution, its depth, its complexity. Also mention that it only takes a few seconds to find these solutions.
\item Going beyond depth 6 (or sometimes even completing depth 6) is computationally infeasible with the current implementation, but depth at most 5 can be searched within at most a few seconds.
\item Combine the tikz and the algebraic notation into one, so I don't need to have two versions. But how? Maybe associate a symbol with each type and only to the types that I use?
\item Partial bijection is the same as partial injection.
\item the exponential solutions can be computed in quadratic time with dynamic programming!
\item Can't explain how formulas are translated into (my definition of) clauses without explaining Skolemization, which is out of scope.
\item Note that in some cases different descriptions of the same problem lead to different solutions (with different complexities).
\item Maybe we are actually guaranteed that the solution is always polynomial-time.
\end{itemize}

Let $p$ be a predicate of arity two s.t. the first argument is associated with domain $a$, and the second argument is associated with domain $b$ (i.e., $p$ represents a relation between sets $a$ and $b$). Then, to restrict all relations representable by $p$ to just functions from $a$ to $b$, in first-order logic one might write
\begin{gather}
  \forall X \in a. \forall Y \in b. \forall Z \in b. p(X, Y) \land p(X, Z) \implies Y = Z \label{eq:def1} \\
  \forall X \in a. \exists Y \in b. p(X, Y). \label{eq:def2}
\end{gather}
The former says that one element of $a$ can map to at \emph{most} one element of $b$, and the latter says that each element of $a$ must map to at \emph{least} one element of $b$. One might add
\begin{equation} \label{eq:injectivity}
  \forall W \in a. \forall X \in a. \forall Y \in b. p(W, Y) \land p(X, Y) \implies W = X
\end{equation}
to restrict $p$ to injections or
\begin{equation}
  \forall Y \in b. \exists X \in a. p(X, Y)
\end{equation}
to ensure surjectivity or remove \cref{eq:def2} to consider partial functions. Lastly, one can replace all occurrences of $b$ with $a$ so as to model endofunctions instead.

\paragraph{Notes.}
\begin{itemize}
\item \textsc{ForcLift} fails on all of these.
\item Functions, surjections, and their partial counterparts are/were already liftable. It seems like lifting injectivity (which is a fairly general property) is the main accomplishment. (But this is just the one I noticed. There may be many others as well.)
\item Here, $[\cdot]$ is the Iverson bracket.
\end{itemize}

\paragraph{Results.}
\begin{itemize}
\item 1d bijections and 1d injections (note that it's the same problem). Depth 3 solution:
  \begin{align*}
    f(n) &= \sum_{m=0}^n \binom{n}{m} (-1)^{n-m}g(n, m) \\
    g(n, m) &= \sum_{l=0}^n \binom{n}{l}[l < 2]g(n-l, m-1) \\
    &= g(n, m - 1) + ng(n - 1, m - 1),
  \end{align*}
  which works with base case $g(n, 0) = 1$.
\item 1d partial injections. 2 solutions at depth 6, but they're too complicated to check by hand. A contradiction with $X \ne x$ constraints makes things complicated.
\item 2d bijections. Depth 3:
  \begin{align*}
    f(m, n) &= \sum_{l=0}^m \binom{m}{l} [l < 2] (1 - [l < 1])f(m-l, n-1) \\
    &= mf(m-1, n-1),
  \end{align*}
  which works with base cases $f(0, 0) = 1$, $f(0, n) = 0$, $f(m, 0) = 0$.
\item 2d injections. Depth 2:
  \begin{align*}
    f(m, n) &= \sum_{l=0}^m \binom{m}{l}[l<2]f(m-l, n-1) \\
    &= f(m, n-1) + mf(m-1, n-1),
  \end{align*}
  which works with base cases $f(0, 0) = 1$ and $f(m, 0) = 0$.
\item 2d partial injections, depth 2. Exactly the same circuit as above but with base case $f(m, 0) = 1$.
\end{itemize}
